---
title: "playground"
format: pdf
execute:
  echo: false       
  warning: false
  message: false
  output: false
---

# Picking a Visualization Angle

There are a number of things we can choose to communicate:

1. Monthly/seasonal patterns of frog observations
2. Amount of frogs in each species observed
3. The range (distance) of frog observations
4. Regions where frogs were observed
5. When observations were made
6. How rare/abundant each frog is


And given that the data spans only 2023, we can use:

1. Monthly or seasonal patterns
2. Range of observations
3. Amount of observations
4. Regions of observations
5. When are the most observations made
6. Rarity

I would like to show rarities, and monthly or seasonal patterns on a map (range) for frog species


# Exploration & Wrangling

```{r}
#| echo: false
#| message: false
#| warning: false

library("tidytuesdayR")
library("tidyverse")
library("dplyr")
library("ggplot2")
library("lubridate")
library("patchwork") # for composing plots
library("showtext") # for fonts
library("ggstream")

tuesdata <- tidytuesdayR::tt_load('2025-09-02')
```

## Supplying Common Name for all Observations


```{r}
#| echo: false
#| message: false
#| warning: false

id_data <- tuesdata$frogID_data
frog_names <- tuesdata$frog_names
class(frog_names)
# class(id_data)
# frog_names$scientificName[frog_names$scientificName == "Heleioporus australiacus australiacus"] <- "Heleioporus australiacus"

# frog_data <- left_join(
#   id_data,
#   frog_names,
#   by = "scientificName",
#   relationship = "many-to-many"
# )

frog_data <- inner_join(
  id_data,
  frog_names,
  by = "scientificName",
  relationship = "many-to-many"
)

# id_data <- id_data %>%
#   mutate(genus_species = str_extract(scientificName, "^\\S+\\s+\\S+"))
# 
# frog_names <- frog_names %>%
#   mutate(genus_species = str_extract(scientificName, "^\\S+\\s+\\S+"))
# 
# # Collapse common names across subspecies
# frog_names_collapsed <- frog_names %>%
#   group_by(genus_species) %>%
#   summarise(
#     commonName = paste(unique(commonName), collapse = " / "),
#     .groups = "drop"
#   )
# 
# # Join back
# frog_data <- left_join(id_data, frog_names_collapsed, by = "genus_species")

clean_frog <- frog_data %>% drop_na()
```

I have found a solution in the chunk above, where for frogs which have been recorded only at the species level, I combine their names in the common name column. So we have things like "Alpine Tree Frog / Whistling Tree Frog".

## Splitting Event Date into Month and Year

```{r}
clean_frog$month <- month(clean_frog$eventDate, label = TRUE, abbr = TRUE)
clean_frog$year <- year(clean_frog$eventDate)

str(clean_frog)
```




## Correcting Column Data Types

I want to create a new column called "Species", taken from the last word in each observation's Scientific Name.
OR
Factor the commonNames column

In addition to correcting the dtypes of other columns

```{r}
#| echo: false
#| message: false
#| warning: false

str(clean_frog)

clean_frog <- clean_frog %>%
  mutate(
    scientificName = as.factor(scientificName),
    stateProvince = as.factor(stateProvince),
    commonName = as.factor(commonName)
    )

str(clean_frog)
```

## Exploratory Visualizations

Since we are visualizing monthly/seasonal variations, rarity, and range, I will visualize:

1. The monthly variations using the commonNames and a faceted bar chart. I would need to group by month (extracted from `eventDate`)
2. The rarity using all the the first and last 5 commonNames, and a bar chart
3. The range using by calculating the minimum and maximum longitudes and latitudes to construct a square whose area I can calculate.

### Visualizing Top and Bottom 5 across months

```{r}
#| execute: false
#| cache: true
#| message: false
#| warning: false

# Exploratory Viz 2, for January
# First, I need to isolate top and bottom 5

top_bottom_frogs <- clean_frog %>%
  group_by(month, commonName) %>%
  summarise(observations = n(), .groups = "drop_last") %>%
  mutate(rank = dense_rank(desc(observations))) %>%
  group_by(month) %>%
  filter(rank <= 5 | rank > max(rank) - 5) %>%
  ungroup()

str(top_bottom_frogs)


# Next, visualize
top_bottom_frogs %>%
  filter(month == "Jan") %>%
  mutate(commonName = factor(commonName)) %>%
ggplot( aes(x = fct_drop(reorder(commonName, observations)),
                         y = observations,
                         fill = rank <= 5)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "grey70"),
                    labels = c("Bottom 5", "Top 5"),
                    name = "Group") +
  labs(
    title = "Top and Bottom 5 Observed Frog Species in January obs",
    x = "Frog Species",
    y = "Number of Observations"
  ) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 7))

```


# NEW PLAN

To ensure that I don't get stuck on irrelevant things, I am going to pick one of the questions on the readme page, and try to visualize it well.
I am going with: Do different frog species have distinct calling seasons?

For this, I will visualize all the frog observations in 12 australian maps (representing the months of the recording year), and point out the most prominent frog in each month. 

Additionally, I will highlight the overall most abundant and rare frog in the entire frog list.

## How do we plot maps in R?

Good question.

I have found that ggplot is able to do so!

From the tutorial I saw:

```{r}
library(giscoR) # This contains map polygons

aussie <- gisco_get_countries(country = "Australia", resolution = 1) ## This extracts Australia
```

Now, our dataset already contains lat and longs so we can plot right away:

```{r}
ggplot() +
  geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
  geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude)) +
  theme_void()
```

I've definitely seen worse plots.

Now, let's attempt to facet it by month:

```{r}
ggplot() +
  geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
  geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude)) +
  facet_wrap(~month, ncol = 4, nrow = 3) +
  theme_void()
```
That seems to have worked, but it doesn't look great.

Going back to our visualization questions,


# NEW NEW PLAN

I am going to convert the Genuses of all the frogs into its own factor column. Then, using the one map, I will show the distributions of the Genuses across Australia. I will also add pictures of the most prominent frogs in each region.


```{r}
# The same way we isolated the genus and species

clean_frog <- clean_frog %>%
  mutate(genus = str_extract(scientificName, "^\\S+"))

clean_frog$genus <- as.factor(clean_frog$genus)
clean_frog$tribe <- as.factor(clean_frog$tribe)
clean_frog$subfamily <- as.factor(clean_frog$subfamily)

str(clean_frog)
```

Now, I am going to visualize (using bar plots and the map), the subfamilies, tribes, and genuses of all the frogs in the dataset:

```{r}
library(paletteer)

# counting each subfamily
# clean_frog %>%
#   group_by(subfamily) %>%
#   summarise(count = n())
# 
# 
# # subfamily bar plot
# clean_frog %>%
#   ggplot(aes(x = subfamily)) +
#   geom_bar() +
#   theme_bw()
# 
# # subfamily map plot
# clean_frog %>%
#   ggplot() +
#   geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
#   geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = subfamily)) +
#   scale_color_brewer(palette = "Set1") +
#   theme_void()
# 
# 
# # counting each tribe
# clean_frog %>%
#   group_by(tribe) %>%
#   summarise(count = n())
# 
# 
# # tribe bar plot
# clean_frog %>%
#   ggplot(aes(x = tribe)) +
#   geom_bar() +
#   theme_bw()
# 
# # tribe map plot
# clean_frog %>%
#   ggplot() +
#   geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
#   geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = tribe)) +
#   scale_color_brewer(palette = "Dark2") +
#   theme_void()

# # counting each genus
# clean_frog %>%
#   group_by(genus) %>%
#   summarise(count = n())
# 
# 
# # genus bar plot
# clean_frog %>%
#   ggplot(aes(x = genus)) +
#   geom_bar() +
#   theme_bw()
# 
# # genus map plot
# clean_frog %>%
#   ggplot() +
#   geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
#   geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = genus)) +
#   theme_void()



```


Seeing the plots, I think the Genus level plots are too busy. I will focus on tribe and subfamilies.
I just need to find color palettes that don't look awful on the plot.

I am going to try to create separate map heatmap (or choropleth) plots of each tribe or subfamily.

```{r}

# subfamily map facet plot
# clean_frog %>%
#   ggplot() +
#   geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
#   geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = subfamily)) +
#   facet_wrap(~subfamily) +
#   scale_color_brewer(palette = "Set1") +
#   theme_void()
# 
# # tribe map facet plot
# clean_frog %>%
#   ggplot() +
#   geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
#   geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = tribe)) +
#   facet_wrap(~tribe) +
#   scale_color_brewer(palette = "Set1") +
#   theme_void()

```

I think I like the tribe plot.

Since we also asked "Do different frog species have distinct calling seasons?", I would like to show the temporal distribution of observations within each tribe across the year, while making sure that the plot is not cluttered.

I can have 6 columns, and within each one, show the scattering of frogs at the four season of the year.

There will be 24 maps total. Do I want that?

If it is cluttered, I can instead add a single histogram to each map plot to show the frequency of observations throughout the year.

```{r}

# Creating a new season column
clean_frog$month_num <- match(clean_frog$month, month.abb)

# clean_frog <- clean_frog %>% 
#   mutate(season = case_when(
#   month_num %in% c(12, 1, 2) ~ "Winter",
#   month_num %in% c(3, 4, 5) ~ "Spring",
#   month_num %in% c(6, 7, 8) ~ "Summer",
#   month_num %in% c(9, 10, 11) ~ "Fall"
# ))

clean_frog <- clean_frog %>%
  mutate(season = cut(month_num,
                    breaks = c(0, 2, 5, 8, 11, 12),
                    labels = c("Winter", "Spring", "Summer", "Fall", "Winter"),
                    include.lowest = TRUE))

clean_frog <- clean_frog %>%
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))

# tribe map facet plot
# clean_frog %>%
#   ggplot() +
#   geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
#   geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = tribe)) +
#   facet_grid(season ~ tribe) +
#   scale_color_brewer(palette = "Set1") +
#   theme_void()
```


So the multiple maps is too busy. Let's see the histograms alongside the map plots:

```{r}

# tribe map facet plot
# clean_frog %>%
#   ggplot() +
#   geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
#   geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = tribe)) +
#   facet_wrap(~tribe, nrow = 1) +
#   scale_color_brewer(palette = "Set1") +
#   labs(title = "Distribution of Observations by Tribe (Map)") +
#   theme_void()
# 
# # Each tribe's temporal distribution
# clean_frog %>%
#   group_by(tribe) %>%
#   mutate(total_tribe_obs = n()) %>%
#   group_by(tribe, month) %>%
#   summarise(
#     monthly_count = n(),
#     total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
#     .groups = 'drop_last' # Keep the grouping structure for the next step
#   ) %>%
#   mutate(proportion = monthly_count / total_tribe_obs) %>%
#   ggplot(aes(x = month, y = proportion, fill = tribe)) +  # Replace 'body_length' with your chosen numeric column
#   geom_col(color = "black", show.legend = FALSE) + # Bins control the smoothness
#   facet_wrap(~tribe, nrow = 1)
#   labs(title = "Distribution of Frog Body Length Observations",
#        x = "Month",
#        y = "Frequency (Number of Observations)") +
#   scale_y_continuous(labels = scales::percent) +
#   scale_color_brewer(palette = "Set1") +
#   theme_void()
```

```{r}
#| cache: false

# tribe map facet plot
map_plot <- clean_frog %>%
  ggplot() +
  geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
  geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = tribe)) +
  facet_wrap(~tribe, nrow = 2, ncol = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Distribution of Observations by Tribe (Map)") +
  theme_void() +
  theme(legend.position = "none")

# Each tribe's temporal distribution
col_plot <- clean_frog %>%
  group_by(tribe) %>%
  mutate(total_tribe_obs = n()) %>%
  group_by(tribe, month) %>%
  summarise(
    monthly_count = n(),
    total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
    .groups = 'drop_last' # Keep the grouping structure for the next step
  ) %>%
  mutate(proportion = monthly_count / total_tribe_obs) %>%
  ggplot(aes(x = proportion, y = month, fill = tribe)) +  # Replace 'body_length' with your chosen numeric column
  geom_col(color = "black", show.legend = FALSE) + # Bins control the smoothness
  facet_wrap(~tribe, nrow = 1) +
  labs(x = "Month",
       y = "Frequency") +
  scale_x_continuous(labels = scales::percent) +
  scale_y_discrete(limits = rev)+
  scale_fill_brewer(palette = "Set1") +
  # theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.text.y = element_blank(),
    
    panel.grid = element_blank(),
    
    panel.background = element_blank(),
    panel.border = element_blank(),

  )
  
map_plot / col_plot #+ plot_layout(heights = c(5, 2))
```

```{r}
tribes <- unique(clean_frog$tribe)

# Create a list of (map + col) pairs for each tribe
plots <- lapply(tribes, function(t) {
  df_t <- clean_frog %>% filter(tribe == t)
  
  # Map for tribe t
  map_t <- ggplot() +
    geom_sf(data = aussie, fill = "grey", alpha = 0.5) +
    geom_point(data = df_t, aes(x = decimalLongitude, y = decimalLatitude), color = "steelblue") +
    labs(title = paste(t)) +
    theme_void()
  
  # Column (temporal) plot for tribe t
  col_t <- df_t %>%
    count(month) %>%
    mutate(proportion = n / sum(n)) %>%
    ggplot(aes(x = proportion, y = month, fill = t)) +
    geom_col(show.legend = FALSE, color = "black") +
    scale_x_continuous(labels = scales::percent) +
    scale_y_discrete(limits = rev)+
    theme_bw(base_size = 8) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
  # Combine side by side; make the map take more width
  map_t + col_t + plot_layout(widths = c(3, 1))
})

# Combine all tribe panels into a 2x3 grid
wrap_plots(plots, ncol = 3)
```

# Hello Again

I am not going to fuss around with patchwork anymore until I can understand it better. Henceforth, I am going to export the individual plots as svg, and do the finishing touches in Inkscape.

```{r}
clean_frog %>%
  ggplot() +
  geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
  geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = tribe)) +
  facet_wrap(~tribe, nrow = 2, ncol = 3) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Distribution of Observations by Tribe (Map)") +
  theme_void() +
  theme(legend.position = "none")
```
```{r}
# clean_frog %>%
#   group_by(tribe) %>%
#   mutate(total_tribe_obs = n()) %>%
#   group_by(tribe, month) %>%
#   summarise(
#     monthly_count = n(),
#     total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
#     .groups = 'drop_last' # Keep the grouping structure for the next step
#   ) %>%
#   mutate(proportion = monthly_count / total_tribe_obs) %>%
#   ggplot(aes(x = proportion, y = month, fill = tribe)) +  # Replace 'body_length' with your chosen numeric column
#   geom_col(color = "black", show.legend = FALSE) + # Bins control the smoothness
#   facet_wrap(~tribe, nrow = 1) +
#   labs(x = "Frequency",
#        y = "Month") +
#   scale_x_continuous(labels = scales::percent) +
#   scale_y_discrete(limits = rev)+
#   scale_fill_brewer(palette = "Set1") +
#   # theme_bw() +
#   theme(
#     plot.title = element_text(hjust = 0.5),
#     strip.background = element_blank(),
#     strip.text.y = element_blank(),
#     
#     panel.grid = element_blank(),
#     
#     panel.background = element_blank(),
#     panel.border = element_blank(),
# 
#   )
```


```{r}
library(RColorBrewer)

clean_frog$tribe <- factor(clean_frog$tribe)  # ensures consistent ordering
tribe_colors <- setNames(brewer.pal(n = length(levels(clean_frog$tribe)), "Set1"),
                         levels(clean_frog$tribe))
```

```{r}

# for (t in levels(clean_frog$tribe)) {
#   df_t <- clean_frog %>% filter(tribe == t)
#   
#   # Map
#   map_t <- ggplot() +
#     geom_sf(data = aussie, fill = "grey", alpha = 0.5) +
#     geom_point(data = df_t, aes(x = decimalLongitude, y = decimalLatitude),
#                color = tribe_colors[t], alpha = 0.7) +
#     labs(title = paste(t, "Map")) +
#     theme_void()
#   
#   # Column plot
#   col_t <- df_t %>%
#     count(month) %>%
#     mutate(proportion = n / sum(n)) %>%
#     ggplot(aes(x = proportion, y = month)) +
#     geom_col(fill = tribe_colors[t], color = "black") +
#     scale_x_continuous(labels = scales::percent) +
#     scale_y_discrete(limits = rev) +
#     labs(title = paste(t, "Monthly Distribution"),
#          x = "Frequency",
#          y = "Month") +
#     theme_bw() +
#     theme(
#       panel.grid = element_blank(),
#       axis.text.x = element_text(angle = 45, hjust = 1)
#     )
  
  # Save both
  # ggsave(filename = paste0("map_", t, ".png"), plot = map_t, width = 6, height = 5, dpi = 300)
  # ggsave(filename = paste0("col_", t, ".png"), plot = col_t, width = 3, height = 5, dpi = 300)
# }
```



## Testing with just two

```{r}
#| output: true

font_add_google(
  "Yusei Magic",
  family = "title_font"
)

showtext_auto()

for (t in levels(clean_frog$tribe)) {
  df_t <- clean_frog %>% filter(tribe == t)
  
  # Map
  map_t <- ggplot() +
    geom_sf(data = aussie, fill = "grey", alpha = 0.5) +
    geom_point(data = df_t, aes(x = decimalLongitude, y = decimalLatitude),
               color = tribe_colors[t], alpha = 0.7) +
    # labs(title = paste(t, "Map")) +
    theme_void()
  
  # Column plot
  col_t <- df_t %>%
    count(month) %>%
    mutate(proportion = n / sum(n)) %>%
    ggplot(aes(x = proportion, y = month)) +
    geom_col(fill = tribe_colors[t], color = "black") +
    scale_x_continuous(labels = scales::percent, expand = c(0, 0)) +
    scale_y_discrete(limits = rev) +
    labs(#title = paste(t, "Monthly Distribution"),
         x = "Frequency",
         y = "Month") +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      panel.border = element_blank(),
      # axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_text(family = "title_font"),
      axis.text = element_text(family = "title_font"),
    )

  col_invert <- df_t %>%
    count(month) %>%
    mutate(proportion = n / sum(n)) %>%
    ggplot(aes(x = month, y = proportion)) +
    geom_col(fill = tribe_colors[t], color = "black") +
    scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
    labs(#title = paste(t, "Monthly Distribution"),
         x = "Month",
         y = "Frequency") +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      panel.border = element_blank(),
      # axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_text(family = "title_font"),
      axis.text = element_text(family = "title_font"),
    )
  
  # Print both
  print(map_t)
  print(col_t)
  print(col_invert)
  
  map_t | col_t
  
  break
}
```




# NOW, HOW ABOUT A STREAM GRAPH?

```{r}
library(devtools)
devtools::install_github("hrbrmstr/streamgraph")
devtools::session_info()

library(streamgraph)


pp <- clean_frog %>%
  mutate(month_date = floor_date(eventDate, "month")) %>%
  group_by(tribe) %>%
  mutate(total_tribe_obs = n()) %>%
  group_by(tribe, month_date) %>%
  summarise(
    monthly_count = n(),
    total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
    .groups = 'drop_last' # Keep the grouping structure for the next step
  ) %>%
  mutate(proportion = monthly_count / total_tribe_obs) %>%
  streamgraph(key = "tribe", value = "proportion", date = "month_date")

pp
```

This isn't bad. Now, I want to:

1. crimp the edges of the plot
2. remove the y-axis ticks
3. match the colours on the stream graphs and the map plots
4. set the global font

```{r}

# Using streamgraph()



# theme_set(theme_minimal(base_family = "Yusei Magic", base_size = 12))

pp <- clean_frog %>%
  mutate(month_date = floor_date(eventDate, "month")) %>%
  group_by(tribe) %>%
  mutate(total_tribe_obs = n()) %>%
  group_by(tribe, month_date) %>%
  summarise(
    monthly_count = n(),
    total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
    .groups = 'drop_last' # Keep the grouping structure for the next step
  ) %>%
  mutate(proportion = monthly_count / total_tribe_obs) %>%
  streamgraph(key = "tribe", value = "proportion", date = "month_date")

pp
```

```{r}
#| output: true

# Using ggstream

library(ggstream)

font_add_google(
  "Yusei Magic",
  family = "title_font"
)

showtext_auto()

palette <- c("")

theme_set(theme_minimal(base_family = "title_font", base_size = 12))

theme_update(
  # plot.title = element_text(
  #   size = 25,
  #   face = "bold",
  #   hjust = .5,
  #   margin = margin(10, 0, 30, 0)
  # ),
  # plot.caption = element_text(
  #   size = 9,
  #   color = "grey40",
  #   hjust = .5,
  #   margin = margin(20, 0, 5, 0)
  # ),
  axis.text.y = element_blank(),
  axis.title = element_blank(),
  # plot.background = element_rect(fill = "grey88", color = NA),
  # panel.background = element_rect(fill = NA, color = NA),
  panel.grid = element_blank(),
  panel.spacing.y = unit(0, "lines"),
  strip.text.y = element_blank(),
  legend.position = "bottom",
  legend.text = element_text(size = 9, color = "grey40"),
  legend.box.margin = margin(t = 30),
  legend.background = element_rect(
    color = "grey40",
    size = .3,
    fill = "grey95"
  ),
  legend.key.height = unit(.25, "lines"),
  legend.key.width = unit(2.5, "lines"),
  # legend.title = "Tribe",
  plot.margin = margin(rep(20, 4))
)

showtext_auto()

gst <- clean_frog %>%
  mutate(month_date = floor_date(eventDate, "month")) %>%
  group_by(tribe) %>%
  mutate(total_tribe_obs = n()) %>%
  group_by(tribe, month_date) %>%
  summarise(
    monthly_count = n(),
    total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
    .groups = 'drop_last' # Keep the grouping structure for the next step
  ) %>%
  mutate(proportion = monthly_count / total_tribe_obs) %>%
  ggplot(
    aes(month_date, proportion,
        color = tribe,
        fill = tribe)
  ) +
  geom_stream(
    geom = "contour",
    color = "white",
    size = 1.25,
    bw = .7 # Controls smoothness
  ) +
  geom_stream(
    geom = "polygon",
    bw = .7,
    size = 0
  ) +
  scale_fill_discrete(name = "Tribe") +     # ← legend title for fill
  scale_color_discrete(name = "Tribe")

gst

```

```{r}
#| output: true

# Using ggstream

font_add_google(
  "Yusei Magic",
  family = "title_font"
)

showtext_auto()

palette <- brewer.pal(n = length(unique(clean_frog$tribe)), name = "Set2")

theme_set(theme_minimal(base_family = "title_font", base_size = 12))

theme_update(
  # plot.title = element_text(
  #   size = 25,
  #   face = "bold",
  #   hjust = .5,
  #   margin = margin(10, 0, 30, 0)
  # ),
  # plot.caption = element_text(
  #   size = 9,
  #   color = "grey40",
  #   hjust = .5,
  #   margin = margin(20, 0, 5, 0)
  # ),
  axis.text.y = element_blank(),
  axis.title = element_blank(),
  # plot.background = element_rect(fill = "grey88", color = NA),
  # panel.background = element_rect(fill = NA, color = NA),
  panel.grid = element_blank(),
  panel.spacing.y = unit(0, "lines"),
  strip.text.y = element_blank(),
  legend.position = "bottom",
  legend.text = element_text(size = 9, color = "grey40"),
  legend.box.margin = margin(t = 30),
  legend.background = element_rect(
    color = "grey40",
    size = .3,
    fill = "grey95"
  ),
  legend.key.height = unit(.25, "lines"),
  legend.key.width = unit(2.5, "lines"),
  # legend.title = "Tribe",
  plot.margin = margin(rep(20, 4))
)

showtext_auto()

gst <- clean_frog %>%
  mutate(month_date = floor_date(eventDate, "month")) %>%
  group_by(tribe) %>%
  mutate(total_tribe_obs = n()) %>%
  group_by(tribe, month_date) %>%
  summarise(
    monthly_count = n(),
    total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
    .groups = 'drop_last' # Keep the grouping structure for the next step
  ) %>%
  mutate(proportion = monthly_count / total_tribe_obs) %>%
  ungroup()

anchors <- gst %>%
  group_by(tribe) %>%
  summarise(
    month_min = min(month_date),
    month_max = max(month_date),
    .groups = "drop"
  ) %>%
  mutate(
    # two points before and after range
    before_1 = month_min - months(2),
    before_2 = month_min - months(1),
    after_1  = month_max + months(1),
    after_2  = month_max + months(2)
  ) %>%
  select(tribe, before_1, before_2, after_1, after_2) %>%
  tidyr::pivot_longer(
    cols = -tribe,
    values_to = "month_date"
  ) %>%
  mutate(
    monthly_count = 0,
    total_tribe_obs = 1,
    proportion = 0
  )

seasons <- as.Date(c(
  "2023-01-01", # Autumn
  "2023-04-01", # Winter
  "2023-07-01", # Spring
  "2023-10-01"  # Summer
))

gst_data_smooth <- bind_rows(gst, anchors)

real_range <- range(gst_data_smooth$month_date)

gst_data_smooth %>%
  mutate(month_date = floor_date(month_date, "month")) %>%
  ggplot(
    aes(x = month_date, y = proportion,
        color = tribe,
        fill = tribe)
  ) +
  geom_stream(
    geom = "contour",
    color = "white",
    size = 1.25,
    bw = .7 # Controls smoothness
  ) +
  geom_stream(
    geom = "polygon",
    bw = .7,
    size = 0
  ) +
  scale_x_date(
    breaks = seasons,
    labels = scales::date_format("%b"),      # short month name (e.g. "Mar")
    # expand = c(0, 0)                         # optional: remove extra space
  ) +
  coord_cartesian(xlim = real_range) +
  scale_fill_manual(values = palette, name = "Tribe") +     # ← legend title for fill
  scale_color_manual(values = palette, name = "Tribe")

```

