---
title: "playground"
format: pdf
execute:
  echo: false       
  warning: false
  message: false
  output: false
---

# Exploration & Wrangling

```{r}
#| echo: false
#| message: false
#| warning: false

library("tidytuesdayR")
library("tidyverse")
library("dplyr")
library("ggplot2")
library("lubridate")
library("patchwork") # for composing plots
library("showtext") # for fonts
library("ggstream")
library("paletteer")
library("devtools")
library("streamgraph")
library("RColorBrewer")

tuesdata <- tidytuesdayR::tt_load('2025-09-02')
```

## Supplying Common Name for all Observations


```{r}
#| echo: false
#| message: false
#| warning: false

id_data <- tuesdata$frogID_data
frog_names <- tuesdata$frog_names

frog_data <- inner_join(
  id_data,
  frog_names,
  by = "scientificName",
  relationship = "many-to-many"
)

clean_frog <- frog_data %>% drop_na()
```

## Splitting Event Date into Month and Year

```{r}
clean_frog$month <- month(clean_frog$eventDate, label = TRUE, abbr = TRUE)
clean_frog$year <- year(clean_frog$eventDate)

str(clean_frog)
```


## Correcting Column Data Types


```{r}
#| echo: false
#| message: false
#| warning: false

str(clean_frog)

clean_frog <- clean_frog %>%
  mutate(
    scientificName = as.factor(scientificName),
    stateProvince = as.factor(stateProvince),
    commonName = as.factor(commonName)
    )

str(clean_frog)
```



```{r}
library(giscoR) # This contains map polygons

aussie <- gisco_get_countries(country = "Australia", resolution = 1) ## This extracts Australia
aussie_rg <- gisco_get_countries(country = "Australia", resolution = 1, spatialtype = "RG") ## This extracts Australia with regional boundaries
```

```{r}
# The same way we isolated the genus and species

clean_frog <- clean_frog %>%
  mutate(genus = str_extract(scientificName, "^\\S+"))

clean_frog$genus <- as.factor(clean_frog$genus)
clean_frog$tribe <- as.factor(clean_frog$tribe)
clean_frog$subfamily <- as.factor(clean_frog$subfamily)

str(clean_frog)
```

```{r}
# Creating a new season column
clean_frog$month_num <- match(clean_frog$month, month.abb)
```



## Testing with just two

```{r}
#| output: true

clean_frog$tribe <- factor(clean_frog$tribe)  # ensures consistent ordering
tribe_colors <- setNames(brewer.pal(n = length(levels(clean_frog$tribe)), "Set1"),
                         levels(clean_frog$tribe))

font_add_google(
  "Yusei Magic",
  family = "title_font"
)

showtext_auto()

for (t in levels(clean_frog$tribe)) {
  df_t <- clean_frog %>% filter(tribe == t)
  
  # Map
  map_t <- ggplot() +
    geom_sf(data = aussie, fill = "grey", alpha = 0.5) +
    geom_point(data = df_t, aes(x = decimalLongitude, y = decimalLatitude),
               color = tribe_colors[t], alpha = 0.7) +
    # labs(title = paste(t, "Map")) +
    theme_void()
  
  # Column plot
  col_t <- df_t %>%
    count(month) %>%
    mutate(proportion = n / sum(n)) %>%
    ggplot(aes(x = proportion, y = month)) +
    geom_col(fill = tribe_colors[t], color = "black") +
    scale_x_continuous(labels = scales::percent, expand = c(0, 0)) +
    scale_y_discrete(limits = rev) +
    labs(#title = paste(t, "Monthly Distribution"),
         x = "Frequency",
         y = "Month") +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      panel.border = element_blank(),
      # axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_text(family = "title_font"),
      axis.text = element_text(family = "title_font"),
    )

  col_invert <- df_t %>%
    count(month) %>%
    mutate(proportion = n / sum(n)) %>%
    ggplot(aes(x = month, y = proportion)) +
    geom_col(fill = tribe_colors[t], color = "black") +
    scale_y_continuous(labels = scales::percent, expand = c(0, 0)) +
    labs(#title = paste(t, "Monthly Distribution"),
         x = "Month",
         y = "Frequency") +
    theme_bw() +
    theme(
      panel.grid = element_blank(),
      panel.border = element_blank(),
      # axis.text.x = element_text(angle = 45, hjust = 1),
      axis.title = element_text(family = "title_font"),
      axis.text = element_text(family = "title_font"),
    )
  
  # Print both
  print(map_t)
  print(col_t)
  print(col_invert)
  
  map_t | col_t
  
  break
}
```




# NOW, HOW ABOUT A STREAM GRAPH?

```{r}
library(devtools)
devtools::install_github("hrbrmstr/streamgraph")
# devtools::session_info()

library(streamgraph)
```


```{r}
#| output: true

# Using ggstream

library(ggstream)

font_add_google(
  "Yusei Magic",
  family = "title_font"
)

showtext_auto()

palette <- c("")

theme_set(theme_minimal(base_family = "title_font", base_size = 12))

theme_update(
  # plot.title = element_text(
  #   size = 25,
  #   face = "bold",
  #   hjust = .5,
  #   margin = margin(10, 0, 30, 0)
  # ),
  # plot.caption = element_text(
  #   size = 9,
  #   color = "grey40",
  #   hjust = .5,
  #   margin = margin(20, 0, 5, 0)
  # ),
  axis.text.y = element_blank(),
  axis.title = element_blank(),
  # plot.background = element_rect(fill = "grey88", color = NA),
  # panel.background = element_rect(fill = NA, color = NA),
  panel.grid = element_blank(),
  panel.spacing.y = unit(0, "lines"),
  strip.text.y = element_blank(),
  legend.position = "bottom",
  legend.text = element_text(size = 9, color = "grey40"),
  legend.box.margin = margin(t = 30),
  legend.background = element_rect(
    color = "grey40",
    size = .3,
    fill = "grey95"
  ),
  legend.key.height = unit(.25, "lines"),
  legend.key.width = unit(2.5, "lines"),
  # legend.title = "Tribe",
  plot.margin = margin(rep(20, 4))
)

showtext_auto()

gst <- clean_frog %>%
  mutate(month_date = floor_date(eventDate, "month")) %>%
  group_by(tribe) %>%
  mutate(total_tribe_obs = n()) %>%
  group_by(tribe, month_date) %>%
  summarise(
    monthly_count = n(),
    total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
    .groups = 'drop_last' # Keep the grouping structure for the next step
  ) %>%
  mutate(proportion = monthly_count / total_tribe_obs) %>%
  ggplot(
    aes(month_date, proportion,
        color = tribe,
        fill = tribe)
  ) +
  geom_stream(
    geom = "contour",
    color = "white",
    size = 1.25,
    bw = .7 # Controls smoothness
  ) +
  geom_stream(
    geom = "polygon",
    bw = .7,
    size = 0
  ) +
  scale_fill_discrete(name = "Tribe") +     # ← legend title for fill
  scale_color_discrete(name = "Tribe")

# gst

```

```{r}
#| output: true

# Using ggstream

font_add_google(
  "Yusei Magic",
  family = "title_font"
)

showtext_auto()

frog_palette <- c("#f0b41a", "#057cc5", "#d7bb71", "#0a2463", "#d8315b", "#1e2f23")

theme_set(theme_minimal(base_family = "title_font", base_size = 12))

theme_update(
  # plot.title = element_text(
  #   size = 25,
  #   face = "bold",
  #   hjust = .5,
  #   margin = margin(10, 0, 30, 0)
  # ),
  # plot.caption = element_text(
  #   size = 9,
  #   color = "grey40",
  #   hjust = .5,
  #   margin = margin(20, 0, 5, 0)
  # ),
  axis.text.y = element_blank(),
  axis.title = element_blank(),
  # plot.background = element_rect(fill = "grey88", color = NA),
  # panel.background = element_rect(fill = NA, color = NA),
  panel.grid = element_blank(),
  panel.spacing.y = unit(0, "lines"),
  strip.text.y = element_blank(),
  legend.position = "bottom",
  legend.text = element_text(size = 9, color = "grey40"),
  legend.box.margin = margin(t = 30),
  legend.background = element_rect(
    color = "grey40",
    size = .3,
    fill = "grey95"
  ),
  legend.key.height = unit(.25, "lines"),
  legend.key.width = unit(2.5, "lines"),
  # legend.title = "Tribe",
  plot.margin = margin(rep(20, 4))
)

showtext_auto()

gst <- clean_frog %>%
  mutate(month_date = floor_date(eventDate, "month")) %>%
  group_by(tribe) %>%
  mutate(total_tribe_obs = n()) %>%
  group_by(tribe, month_date) %>%
  summarise(
    monthly_count = n(),
    total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
    .groups = 'drop_last' # Keep the grouping structure for the next step
  ) %>%
  mutate(proportion = monthly_count / total_tribe_obs) %>%
  ungroup()

anchors <- gst %>%
  group_by(tribe) %>%
  summarise(
    month_min = min(month_date),
    month_max = max(month_date),
    .groups = "drop"
  ) %>%
  mutate(
    # two points before and after range
    before_1 = month_min - months(2),
    before_2 = month_min - months(1),
    after_1  = month_max + months(1),
    after_2  = month_max + months(2)
  ) %>%
  select(tribe, before_1, before_2, after_1, after_2) %>%
  tidyr::pivot_longer(
    cols = -tribe,
    values_to = "month_date"
  ) %>%
  mutate(
    monthly_count = 0,
    total_tribe_obs = 1,
    proportion = 0
  )

seasons <- as.Date(c(
  "2023-01-01", # Autumn
  "2023-04-01", # Winter
  "2023-07-01", # Spring
  "2023-10-01"  # Summer
))

gst_data_smooth <- bind_rows(gst, anchors)

real_range <- range(gst_data_smooth$month_date)

stream_graph <- gst_data_smooth %>%
  mutate(month_date = floor_date(month_date, "month")) %>%
  ggplot(
    aes(x = month_date, y = proportion,
        color = tribe,
        fill = tribe)
  ) +
  geom_stream(
    geom = "contour",
    color = "white",
    size = 1.25,
    bw = .7 # Controls smoothness
  ) +
  geom_stream(
    geom = "polygon",
    bw = .7,
    size = 0
  ) +
  scale_x_date(
    breaks = seasons,
    labels = scales::date_format("%b"),      # short month name (e.g. "Mar")
    # expand = c(0, 0)                         # optional: remove extra space
  ) +
  coord_cartesian(xlim = real_range) +
  scale_fill_manual(values = frog_palette, name = "Tribe") +     # ← legend title for fill
  scale_color_manual(values = frog_palette, name = "Tribe") +
  theme(legend.position = "none")

map_plots <- clean_frog %>%
  ggplot() +
  geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
  geom_point(data  = clean_frog, aes(x = decimalLongitude, y = decimalLatitude, color = tribe)) +
  # facet_wrap(~tribe, nrow = 2, ncol = 3) +
  facet_wrap(~tribe, nrow = 1) +
  scale_color_manual(values = frog_palette) +
  # labs(title = "Distribution of Observations by Tribe (Map)") +
  theme_void() +
  theme(legend.position = "none")

patch <- map_plots / stream_graph
patch_2 <- stream_graph / map_plots

ggsave("stream_map_graph.png", patch)
ggsave("stream_map_graph_2.png", patch_2)
```
# Now, only the stream

```{r}
font_add_google(
  "Yusei Magic",
  family = "title_font"
)

showtext_auto()

frog_palette <- c("#f0b41a", "#057cc5", "#d7bb71", "#0a2463", "#d8315b", "#1e2f23")

theme_set(theme_minimal(base_family = "title_font", base_size = 12))

theme_update(
  # plot.title = element_text(
  #   size = 25,
  #   face = "bold",
  #   hjust = .5,
  #   margin = margin(10, 0, 30, 0)
  # ),
  # plot.caption = element_text(
  #   size = 9,
  #   color = "grey40",
  #   hjust = .5,
  #   margin = margin(20, 0, 5, 0)
  # ),
  axis.text.y = element_blank(),
  axis.title = element_blank(),
  # plot.background = element_rect(fill = "grey88", color = NA),
  # panel.background = element_rect(fill = NA, color = NA),
  panel.grid = element_blank(),
  panel.spacing.y = unit(0, "lines"),
  strip.text.y = element_blank(),
  legend.position = "bottom",
  legend.text = element_text(size = 9, color = "grey40"),
  legend.box.margin = margin(t = 30),
  legend.background = element_rect(
    color = "grey40",
    size = .3,
    fill = "grey95"
  ),
  legend.key.height = unit(.25, "lines"),
  legend.key.width = unit(2.5, "lines"),
  # legend.title = "Tribe",
  plot.margin = margin(rep(20, 4))
)

showtext_auto()

gst <- clean_frog %>%
  mutate(month_date = floor_date(eventDate, "month")) %>%
  group_by(tribe) %>%
  mutate(total_tribe_obs = n()) %>%
  group_by(tribe, month_date) %>%
  summarise(
    monthly_count = n(),
    total_tribe_obs = first(total_tribe_obs), # Get the total count from the group
    .groups = 'drop_last' # Keep the grouping structure for the next step
  ) %>%
  mutate(proportion = monthly_count / total_tribe_obs) %>%
  ungroup()

anchors <- gst %>%
  group_by(tribe) %>%
  summarise(
    month_min = min(month_date),
    month_max = max(month_date),
    .groups = "drop"
  ) %>%
  mutate(
    # two points before and after range
    before_1 = month_min - months(2),
    before_2 = month_min - months(1),
    after_1  = month_max + months(1),
    after_2  = month_max + months(2)
  ) %>%
  select(tribe, before_1, before_2, after_1, after_2) %>%
  tidyr::pivot_longer(
    cols = -tribe,
    values_to = "month_date"
  ) %>%
  mutate(
    monthly_count = 0,
    total_tribe_obs = 1,
    proportion = 0
  )

seasons <- as.Date(c(
  "2023-01-01", # Autumn
  "2023-04-01", # Winter
  "2023-07-01", # Spring
  "2023-10-01"  # Summer
))

gst_data_smooth <- bind_rows(gst, anchors)

real_range <- range(gst_data_smooth$month_date)

gst_data_smooth %>%
  mutate(month_date = floor_date(month_date, "month")) %>%
  ggplot(
    aes(x = month_date, y = proportion,
        color = tribe,
        fill = tribe)
  ) +
  geom_stream(
    geom = "contour",
    color = "white",
    size = 1.25,
    bw = .7 # Controls smoothness
  ) +
  geom_stream(
    geom = "polygon",
    bw = .7,
    size = 0
  ) +
  scale_x_date(
    breaks = seasons,
    labels = scales::date_format("%b"),      # short month name (e.g. "Mar")
    # expand = c(0, 0)                         # optional: remove extra space
  ) +
  coord_cartesian(xlim = real_range) +
  scale_fill_manual(values = frog_palette, name = "Tribe") +     # ← legend title for fill
  scale_color_manual(values = frog_palette, name = "Tribe")


# ggsave("stream_graph.png", patch)
```


# Finally

I think I am going to settle for old faithful: Stacked bar chart, with a log-scaled count axis.

Or parallel coordinates

AND THEN, the map plot

with only the most and least abundant families coloured

There are six categories

```{r}
# Stacked bar chart

clean_frog %>%
  ggplot(aes(x = month, fill = tribe)) +
  geom_bar(position = "stack", color = "black") +
  labs(#title = paste(t, "Monthly Distribution"),
       x = "Month",
       y = "Frequency") +
  theme_bw() +
  scale_y_log10()+
  scale_fill_manual(values = frog_palette)+
  theme(
    panel.grid = element_blank(),
    panel.border = element_blank(),
    # axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(family = "title_font"),
    axis.text = element_text(family = "title_font"),
  )

clean_frog %>%
  group_by(tribe) %>%
  summarise(count = n())
```


# Most and Least Observed Australian Frogs in 2023

```{r}

#e41a1c

clean_frog <- clean_frog %>%
  mutate(season = cut(month_num,
                    breaks = c(0, 2, 5, 8, 11, 12),
                    labels = c("Winter", "Spring", "Summer", "Fall", "Winter"),
                    include.lowest = TRUE))

clean_frog <- clean_frog %>%
  mutate(season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")))

clean_frog <- clean_frog %>%
  mutate(tribe = trimws(tribe))
```


```{r}
ranidae_plots <- clean_frog %>%
  filter(tribe == "Ranidae") %>%
  ggplot() +
  geom_sf(data = aussie_rg, fill = "grey", alpha = 0.5) +
  geom_point(
    aes(x = decimalLongitude, y = decimalLatitude, color = "grey70"),
    alpha = 0.4,
    size = 1.2
  ) +
  geom_point(
    aes(x = decimalLongitude, y = decimalLatitude, color = "#e41a1c"),
    alpha = 0.8,
    size = 2.2
  ) +
  facet_wrap(~season, nrow = 1) +
  labs(title = "Ranidae Observations Through the Seasons") +
  theme_void() +
  theme(legend.position = "none")

ranidae_plots


clean_frog %>%
  ggplot() +
  geom_sf(data = aussie, fill = "grey", alpha = 0.5) + # This is ggplot's function that handles polygon data
  geom_point(aes(x = decimalLongitude, y = decimalLatitude, color = tribe)) +
  facet_wrap(~tribe, nrow = 2, ncol = 3) +
  scale_color_manual(values = frog_palette) +
  labs(title = "Distribution of Observations by Tribe (Map)") +
  theme_void() +
  theme(legend.position = "none")
```




```{r}
# Do you actually have Ranidae and seasonal counts?
table(clean_frog$tribe)
table(clean_frog$season)
table(clean_frog$tribe, clean_frog$season)

# Confirm variable names exist
names(clean_frog)
# Confirm coordinate values look like lon/lat (not reversed)
summary(clean_frog$decimalLongitude); summary(clean_frog$decimalLatitude)

unique(clean_frog$tribe)

```
# I think we were on to something here

```{r}
# Counting the species within each tribe

annotation_data <- clean_frog %>%
  group_by(tribe) %>%
  summarise(
    subspecies = n_distinct(commonName),
    label = paste0(tribe, "\nSubspecies: ", subspecies),
    x = median(decimalLongitude, na.rm = TRUE),
    y = max(decimalLatitude, na.rm = TRUE) + 0.5   # shift upward slightly
  )
```



```{r}
#| output: true

# Using ggstream



font_add_google(
  "Jost",
  family = "title_font"
)

showtext_auto()

frog_palette <- c("#f0b41a", "#057cc5", "#d7bb71", "#0a2463", "#d8315b", "#1e2f23")

theme_set(theme_void(base_family = "title_font", base_size = 12))


showtext_auto()

# map_plots_long <- clean_frog %>%
#   ggplot() +
#   geom_sf(data = aussie, fill = "grey", color = "dimgray", alpha = 0.5, linewidth = 0.2) +
#   geom_point(
#     aes(x = decimalLongitude, y = decimalLatitude, color = tribe),
#     size = 0.5) +
#   facet_wrap(~tribe, nrow = 3, ncol = 2) +
#   scale_color_manual(values = frog_palette) +
#   labs(title = title_text, caption = caption_text) +
#   theme_void(base_family = "title_font")

title_text <- "Diversity of Australia's Frogs"
subtitle_text <- "Australia's frogs belong to 6 tribes, from the widespread Myobatrachidae \nto the north-eastern Ranidae."
caption_text <- "Visualization: Chidiebube Oluoma"

map_plots_wide <- clean_frog %>%
  ggplot() +
  geom_sf(data = aussie, fill = "grey", color = "dimgray", alpha = 0.5, linewidth = 0.2) +
  geom_point(
    aes(x = decimalLongitude, y = decimalLatitude, color = tribe),
    size = 0.5) +
  facet_wrap(~tribe, nrow = 2, ncol = 3) +
  scale_color_manual(values = frog_palette) +
  labs(
    title = title_text,
    subtitle = subtitle_text,
    caption = caption_text
  ) +
  theme_void(base_family = "title_font") +
  theme(
    legend.position = "none",
    plot.title = element_text(
      face = "bold",
      margin = margin(10, 0, 0.5, 0),
      size = 50
    ),
    plot.subtitle = element_text(
      size = 34,
      lineheight = 0.3
    ),
    plot.caption = element_text(
      size = 30,
      lineheight = 0.05,
      margin = margin(0, 0, 5, 0)
    ),
    strip.text = element_text(
      size = 30,
      color = "dimgray"
    ),
    panel.background = element_rect(
      fill = "#f2f2f2"
    ),
    plot.background = element_rect(
      fill = "#f2f2f2"
    )
  )

# map_plots_long
map_plots_wide

# ggsave("map_graph_long.png", map_plots_long)
ggsave("map_graph_wide.png", map_plots_wide)
```


